pirouette: Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-03-18 article.tex
Remove all iffalse sections

\iffalse
%version 1
An optional step is to generate a 'twin tree' 
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree.
The twin tree will be the true tree 
converted to follow the tree prior of the chosen inference model.

To do so we exploit the likelihood function associated with the chosen inference model. We find the parameters (e.g. speciation and extinction rate for a birth-death model) that maximize such likelihood applied to the true tree, conditioned on its number of tips.

We use these parameters to simulate the branching times of the twin tree, 
which we combine with the original tree's topology.

Using a twin tree serves as a control: both the given/true tree
and twin tree will produce an error distribution.
For the true tree, this error distribution is caused by
the shape of the phylogeny (which is different from the tree priors used), 
the stochastic alignment simulation and stochastic Bayesian inference.  
The twin tree is created from the given phylogeny,
as it has the same topology, 
yet its branching times are simulated according to 
the (implemented) generative tree prior.
Due to this, this error distribution for the twin tree is caused 
only by the stochastic alignment simulation and stochastic Bayesian inference.
The difference between these two error distribution is the error caused
by the new/unimplemented tree prior.
\fi

\iffalse
%version 2 with no formulae
An optional step is to generate a 'twin tree' 
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree. We want to produce such tree 'converting' the original tree, created according to the generative prior $\mathit{G}$, into the chosen inference model $\mathit{I}$.

To do so we exploit the likelihood function associated with the chosen inference model $\mathit{I}$. We find the parameters (e.g. speciation and extinction rate for a birth-death model) that maximize such likelihood applied to the true tree, conditioned on its number of tips.

We use these parameters to simulate an equal amount of branching times for the twin tree simulated under the chosen inference model's process. We then combine these branching times with the original tree's topology.

Using a twin tree serves as a control: both the given/true tree and twin tree will produce an error distribution. We assume that these errors are a natural product of the input tree and the intermediate steps of the pipeline. All the factors that can cause such errors are: the branching times of the tree (as a consequence of the tree prior), the topology of the tree, the stochasticity induced by simulating the alignment (step 1a $\rightarrow$ 2a in Fig.~\ref{fig:pipeline}) and the stochasticity induced by the Bayesian inference (step 2a $\rightarrow$ 4a in Fig.~\ref{fig:pipeline}). By imposing the same topology and replicating the pipeline a sufficient number of times we can sort out all the possible causes that are not a direct consequence of the choice of the tree prior.

Potential differences in shapes of error distributions will therefore be consequences of different tree priors for the two parallel pipelines: the use of prior $\mathit{G}$ for the original pipeline and of prior $\mathit{I}$ for the twin pipeline. In the second case, though, such error distribution is not influenced by the mismatch between how the tree is generated and how it is analyzed by BEAST2.
\fi

\giovanni{This is a slightly more formal version}
\richel{As agreed, go ahead and expend this formalization}

%version 3, more formal
An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree, following a parallel pipeline. We want to produce such tree 'converting' the original tree $\tau_{\mathit{P}}$, created according to the generative prior $\mathit{P}$, to be in accordance with the chosen inference model $\mathit{I}$:
\begin{align}
    \tau_{\mathit{P}} = (\Vec{t}_{\mathit{P}}, \psi_{\mathit{P}}) \xrightarrow[]{\mathit{T}} \tau_{\mathit{I}} = (\Vec{t}_{\mathit{I}}, \psi_{\mathit{I}}),
\end{align}
where $\Vec{t}$ are the tree branching times and $\psi$ is tree topology, which provides information on where new lineages branch out at each branching time.

@@

\iffalse
In the previous example, we used a known generative tree prior to match with the inference model.
However, a novel tree prior, by definition, is non-standard, thus cannot be part of a standard inference model (i.e. an inference model using a standard tree prior).
One option is to choose to hand-pick a related 
standard tree prior, as done in ERQs 3 and 5, and combine it with 
a known generative alignment model.
An alternative is to let multiple candidate inference models compete 
and pick the best one, based on the highest evidence (aka marginal likelihood):
when an alignment is simulated from the given tree, one can estimate the
evidence for each inference model given that 
alignment \richel{or the other way around?}.
The benefit of this approach is that this is a commonly used procedure,
which picks an inference model that is
best at explaining the alignment, where more complex inference models are
penalized for having more parameters.
\fi

@@

\iffalse
One option is to choose to hand-pick a related 
standard tree prior, as done in ERQs 3 and 5, and combine it with 
a known generative alignment model.
An alternative is to let multiple candidate inference models compete 
and pick the best one, based on the highest evidence (aka marginal likelihood):
when an alignment is simulated from the given tree, one can estimate the
evidence for each inference model given that 
alignment \richel{or the other way around?}.
The benefit of this approach is that this is a commonly used procedure,
which picks an inference model that is
best at explaining the alignment, where more complex inference models are
penalized for having more parameters.
\fi
\iffalse
A second research question that \verb;pirouette; can answer, is:
\fi

@@

\iffalse
In the second step we set up our experiments. 
Within this ERQ, each of our experiments contain a different
inference model $I$, but has condition $C$ that will only let $I$
be run in inference only if that inference model $I$ is the 
best candidate inference model, that is, 
it is the inference model with the highest evidence.
Setting up experiments like this is thought to be so common,
that \verb;pirouette; has a simple function to do:
\verb;create_all_experiments; will create 40 experiments,
containing the inference models of all combinations of 4 site 
models, 2 clock models and 5 tree priors.
\fi

>>>>

An optional step is to generate a 'twin tree' $\tau_{I}$
(Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 1b),
that will be analyzed in the same way as the true tree, following a parallel pipeline. We want to produce such tree 'converting' the original tree $\tau_{\mathit{P}}$, created according to the generative prior $\mathit{P}$, to be in accordance with the chosen inference model $\mathit{I}$:
\begin{align}
    \tau_{\mathit{P}} = (\Vec{t}_{\mathit{P}}, \psi_{\mathit{P}}) \xrightarrow[]{\mathit{T}} \tau_{\mathit{I}} = (\Vec{t}_{\mathit{I}}, \psi_{\mathit{I}}),
\end{align}
where $\Vec{t}$ are the tree branching times and $\psi$ is tree topology, which provides information on where new lineages branch out at each branching time.

@@

@@

@@
